\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-03-29 honors_thesis/journal_paper/paper.tex
finished second pass

The purpose of this paper is to consider pso parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least minutes, but probably hours or perhaps days.
Our purpose is to explore the question ``Since pso is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

@@

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the previous position. In such
cases all of the next possible positions can be evaluated at the same time as the
evaluation of the current position.
One of these speculative positions will exactly preserve the behavior of the algorithm
and the others are wasted.
This approach has been applied to 
simulated annealing where the next position depends on (1) the last position, (2) a random number
and (3) the result of a binary test comparing the value of the objective function at the current state with the value of the objective function at the
last position.
In this context the next position can be computed in the case where the binary test is true
and in the case where it is false.
in simulated annealing to compute (1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
The test can be evaluated once all three of these evaluations of the objective function are complete, thus
computing all of the objective function evaluations
needed for two iterations in parallel and all at once~\cite{witte-tpds91}.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
By way of contrast, this approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and \emph{value}
at the current position. In this case there are an infinite number of possible speculative
evaluations.

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
We can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce pso we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard pso would have done.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal $\pers$ qualifier to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
neighbor to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself, in that its own position is considered when updating the
particle's $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

\section{Speculative Evaluation in PSO}
\label{sec:sepso}
\subsection{Overview}

The PSO algorithm can be trivially parallelized by distributing the processing needed for each
particle on a separate processor.
But as we have seen in the introduction, for some
functions, and for large numbers of processors where just adding particles (processors)
reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness appreciably faster.

To address this issue we present here a very different use of additional processors.
Performing more iterations instead of adding more particles (for a fixed number of function evaluations),
previous techniques have not addressed this issue.
In order to perform additional iterations by adding additional processors, we
take inspiration from speculative execution in processors.
We do this by refactoring the standard PSO equations as described here.

@@

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all cases.

@@

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
The update equation is taken exactly from the standard update equations~\eqref{eq:velupdate}
and \eqref{eq:posupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t+1}$ we can re-write
$\specvel_{t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t+1} - \ppos_{t+1})
		\bigr]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
\nonumber
	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}

Lastly, we use the normal notation for the value of the objective function at these points,
once again using the case $\casexn$ as an example:
\begin{align}
\nonumber
	\ofunc(\specpos_{t+1}^{\casexn} & (\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t}))= \\
\label{eq:deffcasexn}
	& =  \ofunc(\specpos_{t+1}^{\casexn})
\end{align}

With this notation we can re-write the ``next iteration'' form of the original PSO motion equations
given in equations~\eqref{eq:velupdate}
and \eqref{eq:posupdate} as:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \indic_{t+1}^{c} \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \specvel_{t+1}^{c} \ \indic_{t+1}^{c}
\end{align}
and
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \specpos_{t+1}^{c} \ \indic_{t+1}^{c}
\end{align}

The value of the objective function at $\ppos_{t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{t+1}^{c}) \ \indic_{t+1}^{c}
\end{align}
or, with $\specpos$ and $\indic$ written with the parameters which were omitted above:
\begin{align}
\nonumber
\ofunc & (\ppos_{t+1}) = \sum_{c \in \caseset} \\
\nonumber
& \ofunc(\specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t})) \\
\label{eq:val2updatelong}
& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
\end{align}

In this form the important point to notice is that none of the $\ofunc(\specpos_{t+1}^{c}) \ c \in\ \caseset$ depend upon $f(x_t)$
or any other objective function evaluation at $t+1$.
Thus PSO has been re-factored such that the algorithm can begin computing
all of the objective function evaluations needed in iteration $t+1$ before $f(x_t)$ is computed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
computes $\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
Once all of the evaluations $f(\ppos_{t+1})$ are completed only one
of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
We will refer to the indicator function which is set to one as the
\emph{behavior preserving indicator} or $bp$, since it indicates which of
the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
The rest 
computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
will be ignored, and might just as well never have been computed.
We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
because they may or may not be needed.

The potential value of these speculative children lies in the phrase used above
``this re-factored PSO
can compute $\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
Thus we assume that computing
$\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ is the computationally difficult part of PSO.

Suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. Suppose also that for each particle
the computations
$\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
processors and completed in parallel.
\emph{Given} the values 
of $\ofunc(\ppos_{t+1})$ and all 
$\ofunc(\ppos_{t+1})$ 
the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
is inconsequential relative to the computational cost of obtaining the function values. 
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

While the algebra show that the computation of 
equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate},
There are some issues better seen in an algorithmic setting, in particular the use of the random
vectors, 
$\pURand_{t}$ and $\nURand_{t}$. As with the algebra given above, each transformation of the algorithm is
provably equivalent (except that the algorithms will no longer be able to do an odd number of iterations,
since it will do two at a time). In the algorithmic transformations given below,
instructions are moved around and added, but the
resultant algorithm, produces numerically identical results.

@@

\begin{algorithm}
  \caption{``UpdateBests'': Update personal and neighborhood best values (minimizing)}
  \label{alg:bests}
  \begin{algorithmic}[1]
          \STATE $\pbest_{t}\leftarrow \pbest_{t-1}$
          \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\pbest_{t-1})$
          \IF {$\ofunc(\ppos_{t})<\ofunc(\pbest_{t-1})$}
            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
            \STATE $\fpbest_{t}\leftarrow \fppos_{t}$
          \ENDIF
          \STATE $\nbest_{t}\leftarrow \nbest_{t-1}$
          \STATE $\ofunc(\nbest_{t})\leftarrow \ofunc(\nbest_{t-1})$
          \IF {$\ofunc(\fppos_{t})<\ofunc(\nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\ppos_{t})$
          \ENDIF
          \IF {$\ofunc(\leftn_{t})<\ofunc(nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \leftn_{t}$
            \STATE $\ofunc(pbest_{t})\leftarrow \ofunc(\leftn_{t})$
          \ENDIF
          \IF {$\ofunc(\rightn_{t})<\ofunc(\nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \rightn_{t}$
            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\rightn_{t}$
          \label{alg1:endifs}
          \ENDIF
          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
  \end{algorithmic}
\end{algorithm}

\begin{algorithm} [h!]
  \caption{Standard PSO (minimizing)}
  \label{alg:standard}
  \begin{algorithmic}[1]
    \STATE t=0
    \STATE Initialize the swarm
%    \FOR{$i \in \swarm$}
%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
%    \ENDFOR
    \REPEAT
        \STATE $t=t+1$
        \FOR{$i \in \swarm$}
          \STATE {$\fppos_{t} \leftarrow \ofunc(\ppos_{t})$}
        \ENDFOR
        \FOR{$i \in \swarm$}
          \STATE $\nURand_{t} \Leftarrow RandomVector()$
          \STATE $\nURand_{t} \Leftarrow RandomVector()$
          \STATE UpdateBests
          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
        \ENDFOR
    \UNTIL {$t > max \ iterations$ or convergence criteria met}
  \end{algorithmic}
\end{algorithm}

Using this helper function we can write PSO as shown in~\alg{alg:standard}.
This algorithm can be transformed, while retaining it exact behavior, by
``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s in lines
\ref{alg1:startifs} through
\ref{alg1:endifs} to use the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
positions and objective function value using 
equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
as shown in~\alg{alg:unrolled}

>>>>

The purpose of this paper is to consider PSO parallelization strategies for
clusters of thousands of processors and functions for which a single
evaluation will take at least minutes, but probably hours or perhaps days.
Our purpose is to explore the question ``Since PSO is often run with
a swarm of size of 50-100, what would one do if a few thousand processors were available?''

@@

In still other cases, the next position to be evaluated may depend on the current position and some
discrete valued (binary, for example) test on the value obtained at the current position.
For example, consider a greedy random walk, in which the state only advances if a better
position is found.
That is, if the current value is better than the previous value,
use the sum of the current position and the random number as the next position.
Otherwise, computer the next position using the sum of the previous position and the random number.
In this case the objective function can be evaluated at three positions:
(1) the value of the objective function at the current position,
(2) the value at the position if the test turns out to be true and (3) the value at the position
if the test turns out to be false; all in parallel.
In this example it is important to note that we do not need the value of the objectve function at the current
point to to reduce the infinite number of possible next postions to just two alternatives.
Given the random sample, the next point detments only on the binary test and the
random sample is, by definition, idependent of the positions, and the vaues obtained.
The choice of the next position can be made post-hoc, after the value at the current position is obtained.
For all but the most computationally-trivial objective functions, this approach saves almost half of the
``wall-clock'' time.
This approach has been applied to simulated annealing~\cite{witte-tpds91} which like the greedy random walk,
depends on positions and a binary test.

This approach can not be used in gradient methods where the
step size for sampling typically depends on the gradient and therefor the position and 
a set of infinitly many posible \emph{values} outcomes
obtained at the current position (not binary or even finite).

@@

Furthermore, we show that if we relax the requirements of the algorithm, no
longer demanding that it strictly reproduce the exact behavior of standard PSO, we
can introduce new speculative techniques that often out-perform PSO.
These variants make better use of the extra information obtained from
the extra exploration made by the speculative function evaluations.
We also explore the idea that, like branch prediction in processors, we
need not speculatively evaluate \emph{all} possible future positions,
we can accelerate the algorithm even if we are just \emph{likely} to have guessed right.
By pruning the speculation to just paths that are statistically likely to reproduce the paths that are equivelent to PSO we can also 
speculating several iterations ahead instead of just one.
We also consider several recovery strategies for cases where
the pruned set of speculative evaluations does not contain the evaluation that standard PSO would have done.

@@

At each iteration ($t+1$) of the algorithm, the position $\ppos_t$ and velocity
$\pvel_t$ of each particle are updated as follows:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\label{eq:velupdate}
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\label{eq:posupdate}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
\end{align}
where \( \nURand_{t} \) and \( \nURand_{t} \) are vectors of independent
and identically distributed random numbers drawn from a uniform
distribution, the \( \otimes \) operator is an element-wise vector
multiplication, $\pbest$ (called personal best) is the best position the
current particle has seen, and $\nbest$ (called neighborhood best) is the best
position the neighbors of the current particle have seen~\cite{bratton-sis07}.  \( \ncoeff \), \(
\pcoeff \), and \( \constriction \) are parameters with prescribed values
required to ensure convergence (2.05, 2.05, and .73,
respectively)~\cite{clerc-tec02}. Note also that we use a subscript for the 
time step $t$ or $t+1$ and have moved all other qualifiers,
such as the personal qualifier ``$\pers$'' to the superscript position.
Superscripts can be miss read as exponents, but they are qualifiers throughout this paper.

There are many ways of defining the neighbors of a particle. 
Changing the
neighborhood topology has a significant effect on the performance of the algorithm and 
this effect varies widely by function type, as would be expected by the No Free Lunch
theorem~\cite{wolpert-tec97}.
The two most common topologies used in the
literature are the Ring topology and the Complete
topology~\cite{bratton-sis07}.  In the Ring topology, each particle has one
or more neighbors to either side of it; in the Complete topology, every particle is a
neighbor to every other particle.  In all topologies, a particle is also a
neighbor to itself in that its own position is considered when updating the
particle's personal best, $\nbest$.  Thus with $p$ particles, using the Ring topology each
particle has three neighbors: $i-1$, $i$ (itself), and $i+1$.  With the Complete
topology, each particle has $p$ neighbors.
% TOD USE THIS LATER!
% In this paper we use these
% topologies, as well as a parallel adaptation of the Complete topology, called
% Random, that has been shown to simulate the behavior of Complete with far less
% communication~\cite{mcnabb-cec09}.  In the Random topology, each particle
% randomly picks two other particles to share information with at each iteration,
% along with itself.  Thus in both the Ring and the Random topologies, all
% particles have 3 neighbors.

\section{Speculative Evaluation in PSO}
\label{sec:sepso}

The PSO algorithm can be trivially parallelized by distributing the processing needed for each
particle on a separate processor.
But as we have seen in the introduction, for some
functions, and for large numbers of processors just adding particles
reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness appreciably faster.

To address this issue we present here a very different use of additional processors.
We take inspiration from speculative execution in processors and instead of adding particles
we will employ a speculative evaluation approach like those described in the introduction. 
This allows us to perform iterations two at a time.
To do so we must first re-factor PSO such that the determination of the $\pbest$ and $\nbest$,
which depend upon the value of the objective function, is sepearate from the rest of the computation.

@@

For any topology with 2 neighbors there are 7 possible update cases listed in
Table~\ref{tab:evals}.  This list assumes, as is typically the case, that each
particle is also part of its own neighborhood (can cause an update to
$\nbest$). The list also omits the case where $\nbest$ is updated by
$\ppos_{t}$ but $\pbest$ was not updated, as this is logically impossible.
$\caseset$ represents the set of all update cases.

@@

Again we will drop the parameters for compactness and since
they can be inferred from the subscripts on $\specvel_{t}^{\casexn}$.
The update equation is taken exactly from the standard update equation~\eqref{eq:velupdate}, but
since it is restricted to a specific case, $\casexn$ for this example,
it is no longer necessary to use $\pbest$ and $\nbest$, since in this case
$\pbest_{t+1}=\ppos_{t+1}$ and $\nbest_{t+1}=\nbest_{t}$ we can re-write
$\specvel_{t+1}^{\casexn}$ as:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} &= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t+1}\otimes(\ppos_{t+1} - \ppos_{t+1}) \\
\label{eq:defvcasexn}
			& \quad \quad \quad \; + \ncoeff\nURand_{t+1}\otimes(\nbest_{t} - \ppos_{t+1})
		\bigr]
\end{align}

For each cases identified in
Table~\ref{tab:evals}
and identified by a corresponding indicator variable defined above,
we define also a position update equation with its compact form; once again using $\casexn$ as an example:
\begin{align}
\label{eq:defpcasexn}
	\specpos_{t+1}^{\casexn} & (\ppos_{t}, \pvel_{t},\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t})= \\
\nonumber
	& = \specpos_{t+1}^{\casexn} =  \ppos_{t} + \specvel_{t+1}^{\casexn}
\end{align}
We call these values, $\specpos^{c}) \ c \in\ \caseset$ (7 of them in this case where each particle has 2 neighbors) as ``speculative children''
since they will only be useful if the corresponding $\indic^{c} \ c \in \caseset$ has the value 1.

Lastly, we use the normal notation for the value of the objective function for each speculative child,
once again using the case $\casexn$ as an example:
\begin{align}
\nonumber
	\ofunc(\specpos_{t+1}^{\casexn} & (\ppos_{t+1} ,\leftn_{t+1},\rightn_{t+1} ,\pbest_{t} ,\nbest_{t}))= \\
\label{eq:deffcasexn}
	& =  \ofunc(\specpos_{t+1}^{\casexn})
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate} as:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t}) \\
\nonumber
			& \quad \quad \quad \, + \ncoeff\nURand_{t}\otimes(\nbest_{t} - \ppos_{t})
		\bigr] \indic_{t+1}^{c} \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \specvel_{t+1}^{c} \ \indic_{t+1}^{c}
\end{align}
and the position update 
equation~\eqref{eq:posupdate} as:
\begin{align}
\nonumber
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1} \\
\label{eq:pos2update}
	&= \sum_{c \in \caseset} \ \specpos_{t+1}^{c} \ \indic_{t+1}^{c}
\end{align}
The value of the objective function at $\ppos_{t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \ofunc(\specpos_{t+1}^{c}) \ \indic_{t+1}^{c}
\end{align}

Returning our attention to the computtaion of $\specpos$ in \eqref{eq:pos2update}and writing it with the parameters which were omitted above,
we obtain:
\begin{align}
\nonumber
\ppos_{t+1} &= \sum_{c \in \caseset} \\
\nonumber
& \specpos_{t+1}^{c}(\ppos_{t+1},\leftn_{t+1},\rightn_{t+1},
\pbest_{t},\nbest_{t},\pURand_{t}, \nURand_{t}) \\
\label{eq:val2updatelong}
& \indic_{t}^{c}(\ofunc ( \ppos_{t+1} ) ,\ofunc(\leftn_{t+1}),\ofunc(\rightn_{t+1}) ,\ofunc(\pbest_{t}) ,\ofunc(\nbest_{t}))
\end{align}
In this form the important point to notice is that none of the $\specpos_{t+1}^{c} \ c \in\ \caseset$ depend upon $f(x_t)$
or any other objective function evaluation at $t$.
Thus PSO has been re-factored such that the algorithm can begin computing
all 7 of the objective function evaluations potentially needed in iteration $t+1$ before $f(x_t)$ is computed.

This refactoring comes a significant expense, instead of just computing
$\ofunc(\ppos_{i,t})$ and then $\ofunc(\ppos_{i,t+1})$ in the usual way, this refactored PSO
computes $\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time.
Once the evaluation of $f(\ppos_{t})$ is completed only one
of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ will be set to one.
We will refer to the indicator function which is set to one as the
\emph{behavior preserving indicator} or $bp$, since it indicates which of
the computations $\ofunc(\specpos_{t}^{c}) \  c \in \caseset$ preservers the behavior of PSO.
The rest 
computations $\ofunc(\specpos_{t}^{c}) \ c \in \caseset-\{bp\}$
will be ignored, and might just as well never have been computed.
We call these evaluations of $\ofunc(\specpos_{t+1}^{c})$ ``speculative children''
because they may or may not be needed.

The potential value of these speculative children lies in the phrase used above
``this re-factored PSO
can compute $\ofunc(\ppos_{t+1})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ \ c \in \caseset$ at the same time''. 
Remember that our intent is to accelerate parallel PSO and that we
are interested in objective functions that have at least non-trivial computational
complexity.
That is, we assume that computing
$\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ are the computationally difficult parts.

For example, suppose that unlimited processors were available and that the evaluation of 
the objective function took 1 hour. Suppose also that for each particle
the computations
$\ofunc(\ppos_{t})$ and all 
of the $\ofunc(\specpos_{t+1}^{c}) \ c \in \caseset$ were sent to separate
processors and completed in parallel.
\emph{Given} the values 
of $\ofunc(\ppos_{t})$ for all particles
the computation of the indicator functions $\indic_{t}^{c} \ c \in \caseset$ and to complete iteration t+1
is inconsequential relative to the computational cost of obtaining the function values. 
In this environment all of the function evaluations needed for 2 iterations of PSO
could be completed in just one hour where as the conventional implementation would
require two hours for those same 2 iterations of PSO.

While the algebra show that the computation of 
equations~\eqref{eq:vel2update} and \eqref{eq:pos2update} are identical to the 
equations~\eqref{eq:velupdate} and \eqref{eq:posupdate},
There are some issues better seen in an algorithmic setting, in particular the use of the random
vectors, 
$\pURand_{t}$ and $\nURand_{t}$.
As with the algebra given above, each transformation (used below) of the algorithm is
provably equivalent; except that the algorithms will no longer be able to do an odd number of iterations,
since it will do two at a time. In the algorithmic transformations given below,
instructions are moved around and added, but the
resultant algorithm, produces numerically identical results.

@@

\begin{algorithm}
  \caption{``UpdateBests'': Update personal and neighborhood best values (minimizing)}
  \label{alg:bests}
  \begin{algorithmic}[1]
          \STATE $\pbest_{t}\leftarrow \pbest_{t-1}$
          \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\pbest_{t-1})$
          \IF {$\ofunc(\ppos_{t})<\ofunc(\pbest_{t-1})$}
            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
            \STATE $\fpbest_{t}\leftarrow \fppos_{t}$
          \ENDIF
          \STATE $\nbest_{t}\leftarrow \nbest_{t-1}$
          \STATE $\ofunc(\nbest_{t})\leftarrow \ofunc(\nbest_{t-1})$
          \IF {$\ofunc(\fppos_{t})<\ofunc(\nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \ppos_{t}$
            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\ppos_{t})$
          \ENDIF
          \IF {$\ofunc(\leftn_{t})<\ofunc(nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \leftn_{t}$
            \STATE $\ofunc(pbest_{t})\leftarrow \ofunc(\leftn_{t})$
          \ENDIF
          \IF {$\ofunc(\rightn_{t})<\ofunc(\nbest_{t})$}
            \STATE $\pbest_{t}\leftarrow \rightn_{t}$
            \STATE $\ofunc(\pbest_{t})\leftarrow \ofunc(\rightn_{t}$
          \ENDIF
  \end{algorithmic}
\end{algorithm}

\begin{algorithm} [h!]
  \caption{Standard PSO (minimizing)}
  \label{alg:standard}
  \begin{algorithmic}[1]
    \STATE t=0
    \STATE Initialize the swarm
%    \FOR{$i \in \swarm$}
%      \STATE {$\ppos_{i,t} \leftarrow Random position$}
%      \STATE {$\pvel_{i,t} \leftarrow Random velocities$}
%      \STATE {$\fpbest_{i,t} \leftarrow Max Float$}
%      \STATE {$\fnbest_{i,t} \leftarrow Max Float$}
%    \ENDFOR
    \REPEAT
        \STATE $t=t+1$
        \FOR{each particle in the swarm}
          \STATE {$\fppos_{t} \leftarrow \ofunc(\ppos_{t})$}
        \ENDFOR
        \FOR{each particle in the swarm}
          \STATE $\nURand_{t} \Leftarrow RandomVector()$
          \STATE $\nURand_{t} \Leftarrow RandomVector()$
          \STATE UpdateBests
          \STATE Update $\pvel_{t+1}$ as shown in equation~\eqref{eq:velupdate} 
          \STATE Update $\ppos_{t+1}$ as shown in equation~\eqref{eq:posupdate}
        \ENDFOR
    \UNTIL {$t > max \ iterations$ or convergence criteria met}
  \end{algorithmic}
\end{algorithm}

Using this helper function we can write PSO as shown in~\alg{alg:standard}.
This algorithm can be transformed, while retaining it exact behavior, by
``unrolling'' one iteration of the ``Repeat'' loop, redundantly expressing work done in the ``if'''s
in~\alg{alg:bests}
using the indicator functions $\indic_{t+1}$ and re-writing the computation of the next velocity,
positions and objective function value using 
equations~\eqref{eq:vel2update}, \eqref{eq:pos2update} and \eqref{eq:val2update} 
as shown in the new version of the algorithm as shown in~\alg{alg:unrolled}
